Technical comment on `Negative-assortative mating for color in wolves'

### 2022-03-04 ms/ms.Rmd
small changes to ms

In @hedrick_negative-assortative_2016, the frequency of gray $\times$ black matings is incorrectly written as $2 P (H + Q)$ (*cf* Table 1). However, this value does not account for all possible outcomes that result in Gray $\times$ black matings (Table \@ref(tab:probabilities)). As a result, the genotype frequencies (Table \@ref(tab:genotypes)) do not sum to 1 as they should if they account for all possible outcomes. Ironically, I found an analogous derivation in @hedrick_population_2012 for positive-assortative mating where the genotype frequencies do sum to 1 when ignoring other evolutionary forces such as selection. The correct expression derived by summing all ways gray $\times$ black matings can occur is provided in Table \@ref(tab:genotypes). The corrected genotype frequencies sum to 1 as expected (see code for analytical derivation in Supporting Information).

@@

read_rds("objects/table1_genotypes.rds") |>
  mutate(
    `Parent 1` = glue("${parent1}$"),
    `Pr[Parent 1]` = glue("${prob1_symbol}$"),
    `Pr[Mating]` = glue("${prob_mating_symbol}$"),
    `Parent 2` = glue("${parent2}$"),
    `Pr[Parent 2]` = glue("${prob2_symbol}$"),
    `Pr[Total]` = glue("${genotype_frequency_symbol}$") |>
      str_replace_all("\\*\\*", " ^ ") |> 
      str_replace_all("\\*", " ") 
  ) |>
  select(`Parent 1`, `Pr[Parent 1]`, Mating = mating, `Pr[Mating]`, `Parent 2`, 
         `Pr[Parent 2]`, `Pr[Total]`) |>
  arrange(desc(`Parent 1`), desc(`Parent 2`), Mating) |>
  mutate(Color = case_when(
    `Parent 1` == "$kk$" & `Parent 2` == "$kk$" ~ "Gray $\\times$ gray",
    `Parent 1` == "$kk$" & `Parent 2` == "$K-$" ~ "Gray $\\times$ black",
    `Parent 1` == "$K-$" & `Parent 2` == "$kk$" ~ "Gray $\\times$ black",
    `Parent 1` == "$K-$" & `Parent 2` == "$K-$" ~ "Black $\\times$ black",
  )) |>
  knitr::kable(format = "markdown", caption = "The probability of every mating outcome in the negative-assortative mating model analyzed by Hedrick *et al.* (2016).", align = "llllllll")

@@

In conclusion, the mathematical errors in @hedrick_negative-assortative_2016 do not undermine their primary conclusion that negative-assortative mating by color may explain the distribution of genotype frequencies at the beta definsin locus in the Yellowstone population of wolves (*Canis lupus*). The derivation here may prove useful to future research on detecting evidence for negative-assortative mating.

>>>>

In @hedrick_negative-assortative_2016, the frequency of gray $\times$ black matings is incorrectly written as $2 P (H + Q)$ (*cf* Table 1). However, this value does not account for all possible outcomes that result in gray $\times$ black matings (Table \@ref(tab:probabilities)). As a result, the genotype frequencies (Table \@ref(tab:genotypes)) do not sum to 1 as they should if they account for all possible outcomes. Ironically, I found an analogous derivation in @hedrick_population_2012 for positive-assortative mating where the genotype frequencies do sum to 1 when ignoring other evolutionary forces such as selection. The correct expression derived by summing all ways gray $\times$ black matings can occur is provided in Table \@ref(tab:genotypes). The corrected genotype frequencies sum to 1 as expected (see code for analytical derivation in Supporting Information).

@@

read_rds("objects/table1_genotypes.rds") |>
  mutate(
    `Parent 1` = glue("${parent1}$"),
    `Pr[Parent 1]` = glue("${prob1_symbol}$"),
    `Pr[Mating]` = glue("${prob_mating_symbol}$"),
    `Parent 2` = glue("${parent2}$"),
    `Pr[Parent 2]` = glue("${prob2_symbol}$"),
    `Pr[Total]` = glue("${genotype_frequency_symbol}$") |>
      str_replace_all("\\*\\*", " ^ ") |> 
      str_replace_all("\\*", " ") 
  ) |>
  select(`Parent 1`, `Pr[Parent 1]`, Mating = mating, `Pr[Mating]`, `Parent 2`, 
         `Pr[Parent 2]`, `Pr[Total]`) |>
  arrange(desc(`Parent 1`), desc(`Parent 2`), Mating) |>
  mutate(Color = case_when(
    `Parent 1` == "$kk$" & `Parent 2` == "$kk$" ~ "Gray $\\times$ gray",
    `Parent 1` == "$kk$" & `Parent 2` == "$K-$" ~ "Gray $\\times$ black",
    `Parent 1` == "$K-$" & `Parent 2` == "$kk$" ~ "Gray $\\times$ black",
    `Parent 1` == "$K-$" & `Parent 2` == "$K-$" ~ "Black $\\times$ black",
  )) |>
  knitr::kable(format = "markdown", caption = "The probability of every mating outcome in the negative-assortative mating model analyzed by Hedrick *et al.* (2016). For the notation, the probability of event *X* is Pr[*X*]. The total probabilities for each row are derived from the product of all probabilities in the same row, Pr[Total] = Pr[Parent 1] $\\times$ Pr[Mating] $\\times$ Pr[Parent 2].", align = "llllllll")

@@

In conclusion, the mathematical error in @hedrick_negative-assortative_2016 does not undermine their primary conclusion that negative-assortative mating by color may explain the distribution of genotype frequencies at the beta definsin locus in the Yellowstone population of wolves (*Canis lupus*). The derivation here may prove useful to future research on negative-assortative mating.

### 2022-03-04 ms/ms.tex
small changes to ms

In \protect\hyperlink{ref-hedrick_negative-assortative_2016}{Hedrick et al.} (\protect\hyperlink{ref-hedrick_negative-assortative_2016}{2016}), the frequency of gray \(\times\) black matings is incorrectly written as \(2 P (H + Q)\) (\emph{cf} Table 1). However, this value does not account for all possible outcomes that result in Gray \(\times\) black matings (Table \ref{tab:probabilities}). As a result, the genotype frequencies (Table \ref{tab:genotypes}) do not sum to 1 as they should if they account for all possible outcomes. Ironically, I found an analogous derivation in \protect\hyperlink{ref-hedrick_population_2012}{Hedrick and Ritland} (\protect\hyperlink{ref-hedrick_population_2012}{2012}) for positive-assortative mating where the genotype frequencies do sum to 1 when ignoring other evolutionary forces such as selection. The correct expression derived by summing all ways gray \(\times\) black matings can occur is provided in Table \ref{tab:genotypes}. The corrected genotype frequencies sum to 1 as expected (see code for analytical derivation in Supporting Information).

\begin{longtable}[]{@{}llllllll@{}}
\caption{\label{tab:probabilities}The probability of every mating outcome in the negative-assortative mating model analyzed by Hedrick \emph{et al.} (2016).}\tabularnewline
\toprule
Parent 1 & Pr{[}Parent 1{]} & Mating & Pr{[}Mating{]} & Parent 2 & Pr{[}Parent 2{]} & Pr{[}Total{]} & Color \\
\midrule
\endfirsthead
\toprule
Parent 1 & Pr{[}Parent 1{]} & Mating & Pr{[}Mating{]} & Parent 2 & Pr{[}Parent 2{]} & Pr{[}Total{]} & Color \\
\midrule
\endhead
\(kk\) & \(P\) & assortative & \(A\) & \(kk\) & \(0\) & \(0\) & Gray \(\times\) gray \\
\(kk\) & \(P\) & random & \((1 - A)\) & \(kk\) & \(P\) & \(P ^ 2 (1 - A)\) & Gray \(\times\) gray \\
\(kk\) & \(P\) & assortative & \(A\) & \(K-\) & \(1\) & \(A P\) & Gray \(\times\) black \\
\(kk\) & \(P\) & random & \((1 - A)\) & \(K-\) & \((1 - P)\) & \(P (A - 1) (P - 1)\) & Gray \(\times\) black \\
\(K-\) & \((1 - P)\) & assortative & \(A\) & \(kk\) & \(1\) & \(A (1 - P)\) & Gray \(\times\) black \\
\(K-\) & \((1 - P)\) & random & \((1 - A)\) & \(kk\) & \(P\) & \(P (A - 1) (P - 1)\) & Gray \(\times\) black \\
\(K-\) & \((1 - P)\) & assortative & \(A\) & \(K-\) & \(0\) & \(0\) & Black \(\times\) black \\
\(K-\) & \((1 - P)\) & random & \((1 - A)\) & \(K-\) & \((1 - P)\) & \((1 - A) (P - 1) ^ 2\) & Black \(\times\) black \\
\bottomrule
\end{longtable}

@@

In conclusion, the mathematical errors in \protect\hyperlink{ref-hedrick_negative-assortative_2016}{Hedrick et al.} (\protect\hyperlink{ref-hedrick_negative-assortative_2016}{2016}) do not undermine their primary conclusion that negative-assortative mating by color may explain the distribution of genotype frequencies at the beta definsin locus in the Yellowstone population of wolves (\emph{Canis lupus}). The derivation here may prove useful to future research on detecting evidence for negative-assortative mating.

>>>>

In \protect\hyperlink{ref-hedrick_negative-assortative_2016}{Hedrick et al.} (\protect\hyperlink{ref-hedrick_negative-assortative_2016}{2016}), the frequency of gray \(\times\) black matings is incorrectly written as \(2 P (H + Q)\) (\emph{cf} Table 1). However, this value does not account for all possible outcomes that result in gray \(\times\) black matings (Table \ref{tab:probabilities}). As a result, the genotype frequencies (Table \ref{tab:genotypes}) do not sum to 1 as they should if they account for all possible outcomes. Ironically, I found an analogous derivation in \protect\hyperlink{ref-hedrick_population_2012}{Hedrick and Ritland} (\protect\hyperlink{ref-hedrick_population_2012}{2012}) for positive-assortative mating where the genotype frequencies do sum to 1 when ignoring other evolutionary forces such as selection. The correct expression derived by summing all ways gray \(\times\) black matings can occur is provided in Table \ref{tab:genotypes}. The corrected genotype frequencies sum to 1 as expected (see code for analytical derivation in Supporting Information).

\begin{longtable}[]{@{}llllllll@{}}
\caption{\label{tab:probabilities}The probability of every mating outcome in the negative-assortative mating model analyzed by Hedrick \emph{et al.} (2016). For the notation, the probability of event \emph{X} is Pr{[}\emph{X}{]}. The total probabilities for each row are derived from the product of all probabilities in the same row, Pr{[}Total{]} = Pr{[}Parent 1{]} \(\times\) Pr{[}Mating{]} \(\times\) Pr{[}Parent 2{]}.}\tabularnewline
\toprule
Parent 1 & Pr{[}Parent 1{]} & Mating & Pr{[}Mating{]} & Parent 2 & Pr{[}Parent 2{]} & Pr{[}Total{]} & Color \\
\midrule
\endfirsthead
\toprule
Parent 1 & Pr{[}Parent 1{]} & Mating & Pr{[}Mating{]} & Parent 2 & Pr{[}Parent 2{]} & Pr{[}Total{]} & Color \\
\midrule
\endhead
\(kk\) & \(P\) & assortative & \(A\) & \(kk\) & \(0\) & \(0\) & Gray \(\times\) gray \\
\(kk\) & \(P\) & random & \((1 - A)\) & \(kk\) & \(P\) & \(P ^ 2 (1 - A)\) & Gray \(\times\) gray \\
\(kk\) & \(P\) & assortative & \(A\) & \(K-\) & \(1\) & \(A P\) & Gray \(\times\) black \\
\(kk\) & \(P\) & random & \((1 - A)\) & \(K-\) & \((1 - P)\) & \(P (A - 1) (P - 1)\) & Gray \(\times\) black \\
\(K-\) & \((1 - P)\) & assortative & \(A\) & \(kk\) & \(1\) & \(A (1 - P)\) & Gray \(\times\) black \\
\(K-\) & \((1 - P)\) & random & \((1 - A)\) & \(kk\) & \(P\) & \(P (A - 1) (P - 1)\) & Gray \(\times\) black \\
\(K-\) & \((1 - P)\) & assortative & \(A\) & \(K-\) & \(0\) & \(0\) & Black \(\times\) black \\
\(K-\) & \((1 - P)\) & random & \((1 - A)\) & \(K-\) & \((1 - P)\) & \((1 - A) (P - 1) ^ 2\) & Black \(\times\) black \\
\bottomrule
\end{longtable}

@@

In conclusion, the mathematical error in \protect\hyperlink{ref-hedrick_negative-assortative_2016}{Hedrick et al.} (\protect\hyperlink{ref-hedrick_negative-assortative_2016}{2016}) does not undermine their primary conclusion that negative-assortative mating by color may explain the distribution of genotype frequencies at the beta definsin locus in the Yellowstone population of wolves (\emph{Canis lupus}). The derivation here may prove useful to future research on negative-assortative mating.
